How Drug Usage During Labor and Delivery Affects Infant Health: A Predictive Analysis of APGAR5 Scores

### 2024-11-23 paper/paper.qmd
tables and graphs for data/model results

Some paragraphs about how we go from a phenomena in the world to an entry in the dataset.

@@

# Create heatmap
# Create a mapping of custom labels for drug usage categories
custom_labels <- c(
  indc = "Induction",
  augmt = "Augmentation",
  ster = "Steroids",
  antb = "Antibiotics",
  chor = "Chorioamnionitis",
  anes = "Anesthesia"
)
ggplot(heatmap_long, aes(x = APGAR5, y = Drug_Usage, fill = Count)) +
  geom_tile(color = "white") +
  scale_fill_gradient(
    low = "skyblue", 
    high = "blue", 
    name = "Count of Yes Responses", 
    guide = guide_colorbar(
      name = "Yes Responses",
      barwidth = 1,  # Adjust the width of the legend
      barheight = 6  # Adjust the height of the legend
    )
  ) +
  scale_x_continuous(breaks = 0:10) +  # Show all integers from 0 to 10
  scale_y_discrete(labels = custom_labels) +  # Apply custom labels to y-axis (Drug Usage Categories)
  labs(
    x = "Five-Minute APGAR Score",
    y = "Treatments During Labor and Delivery"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    axis.text.x = element_text(hjust = 1, vjust = 1),  # Angle and align x-axis text
    axis.text.y = element_text(size = 9),  # Increase y-axis text size
    legend.text = element_text(size = 7),  # Reduce legend text size
    legend.title = element_text(size = 7),  # Reduce legend title size
  )

@@

Our results are summarized in @tbl-modelresults.

@@

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

>>>>

The data transforms the phenomenon of infant health status into data, defining measurable variables that can capture the complex factors influencing the APGAR5 score. This transformation is achieved through the collection of information on various treatments and conditions experienced during labor and delivery, such as steroid use, chorioamnionitis, antibiotics, and anesthesia, among others. These variables, which are typically binary, are then recorded through surveys or medical records, using checkboxes or scales that allow for quantification. The APGAR5 score itself, is also a simplified numerical representation of an infant's health, which may overlook the complexities of their condition.

By structuring these variables into numerical or categorical forms, the data can be analyzed and used to model the relationship between treatments and the APGAR5 score. However, this process involves challenges such as measurement error and biases, especially when data collection relies on self-reporting or clinician-recorded information.

@@

# Create heatmap
# Create a mapping of custom labels for drug usage categories
custom_labels <- c(
  indc = "Induction",
  augmt = "Augmentation",
  ster = "Steroids",
  antb = "Antibiotics",
  chor = "Chorioamnionitis",
  anes = "Anesthesia"
)
ggplot(heatmap_long, aes(x = APGAR5, y = Drug_Usage, fill = Count)) +
  geom_tile(color = "white") +
  scale_fill_gradient(
    low = "skyblue", 
    high = "blue", 
    name = "Yes Responses", 
    guide = guide_colorbar(
      name = "Yes Responses",
      barwidth = 1,  # Adjust the width of the legend
      barheight = 6  # Adjust the height of the legend
    )
  ) +
  scale_x_continuous(breaks = 0:10) +  # Show all integers from 0 to 10
  scale_y_discrete(labels = custom_labels) +  # Apply custom labels to y-axis (Drug Usage Categories)
  labs(
    x = "Five-Minute APGAR Score",
    y = "Treatments During Labor and Delivery"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    axis.text.x = element_text(hjust = 1, vjust = 1),  # Angle and align x-axis text
    axis.text.y = element_text(size = 9),  # Increase y-axis text size
    legend.text = element_text(size = 7),  # Reduce legend text size
    legend.title = element_text(size = 7),  # Reduce legend title size
  )

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(randomForest)
library(caret)

rf_model <-
  readRDS(file = here::here("models/random_forest.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-rfsummary1
#| tbl-cap: Variable importance of treatments during labor and delivery on infant health, based on Random Forest model
#| warning: false

# Load necessary libraries
library(tidyverse)
library(caret)
library(knitr)
library(kableExtra)

# Custom labels for the variables
custom_labels <- c(
  "indc1" = "Induction of Labor",
  "augmt1" = "Augmentation of Labor",
  "ster1" = "Steroids",
  "antb1" = "Antibiotics",
  "chor1" = "Chorioamnionitis",
  "anes1" = "Anesthesia"
)

# Extract variable importance using caret
importance_data <- varImp(rf_model, scale = TRUE)

# Convert varImp.train object to a data frame
importance_table <- as.data.frame(importance_data$importance)

# Add variable names as a new column
importance_table$Variable <- rownames(importance_table)

# Transform the data frame
importance_table <- importance_table %>%
  mutate(
    `Treatment Type` = recode(Variable, !!!custom_labels),  # Replace with custom labels
    `Variable Importance` = Overall / sum(Overall) * 100  # Convert to percentages
  ) %>%
  select(`Treatment Type`, `Variable Importance`) %>%  # Select relevant columns
  arrange(match(`Treatment Type`, custom_labels)) %>%  # Sort by custom_labels order
  mutate(`Variable Importance` = round(`Variable Importance`, 2))  # Round to 2 decimal places

# Display the importance table using kableExtra
importance_table %>%
  kable(
    col.names = c("Treatment Type", "Variable Importance (%)")) %>%
  kable_styling(full_width = FALSE, position = "center")

```{r}
#| echo: false
#| eval: true
#| label: fig-rfsumbar
#| fig-cap: Variable importance of treatments during labor and delivery on infant health, expressed in bar plots
#| warning: false

# Load necessary library
library(ggplot2)

# Horizontal bar plot
ggplot(importance_table, aes(x = `Variable Importance`, y = `Treatment Type`)) +
  geom_bar(stat = "identity", fill = "#DDA0DD", color = "grey") +
  labs(
    x = "Variable Importance (%)",
    y = "Treatment Type"
  ) +
  theme_minimal() +
  theme(
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10)
  )

```{r}
#| echo: false
#| eval: true
#| label: tbl-rfsummary2
#| tbl-cap: Model accuracy metrics for Random Forest Model
#| warning: false
#| message: false

# Load necessary libraries
library(tidyverse)
library(caret)
library(knitr)
library(kableExtra)

# Predict on test data
train_data <- read_csv("../data/02-analysis_data/train_test_data/train_data.csv")
test_data <- read_csv("../data/02-analysis_data/train_test_data/test_data.csv")
train_data <- train_data %>%
  mutate(across(c("indc", "augmt", "ster", "antb", "chor", "anes"), as.factor))
test_data <- test_data %>%
  mutate(across(c("indc", "augmt", "ster", "antb", "chor", "anes"), as.factor))
predictions <- predict(rf_model, newdata = test_data)

# Calculate performance metrics
actual <- test_data$apgar5
rmse <- sqrt(mean((predictions - actual)^2))  # RMSE
mae <- mean(abs(predictions - actual))       # MAE
rss <- sum((predictions - actual)^2)         # Residual Sum of Squares
tss <- sum((actual - mean(actual))^2)        # Total Sum of Squares
r_squared <- 1 - (rss / tss)                 # R-Squared
adjusted_r_squared <- 1 - ((1 - r_squared) * ((nrow(test_data) - 1) / (nrow(test_data) - length(rf_model$importance) - 1)))

# Create a summary table
accuracy_table <- tibble(
  Metric = c("RMSE", "MAE", "R-Squared", "Adjusted R-Squared"),
  Value = c(round(rmse, 2), round(mae, 2), round(r_squared, 2), round(adjusted_r_squared, 2))
)

# Display the table
accuracy_table %>%
  kable(
    col.names = c("Metric", "Value")
  ) %>%
  kable_styling(full_width = FALSE, position = "center")

@@
